---
title: "Ontology concepts and tools in Bioconductor for Bioc2024"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Ontology concepts and tools in Bioconductor for Bioc2024}
  %\VignetteEncoding{UTF-8}
bibliography: ontobib.bib
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Objectives

- introduction to ontology-related computations in R with the ORDO rare disease ontology and CL cell ontology

- review of general ontology resources and the OWL RDF/XML serialization

- retrieving, updating, searching for OWL representations of ontologies using `BiocFileCache` and `owl2cache`

- Ontology mapping of phenotypes of the GWAS catalog using
Experimental Feature Ontology

# Introduction

## ORDO for rare diseases

Our first example uses the ORDO ontology, which is distributed
in compressed OWL format
with the ontoProc package.

Here are the first 18 lines of the OWL XML:
```{r lk1, message=FALSE}
library(ontoProc)
pa = get_ordo_owl_path()
readLines(pa, 18)
```
This establishes a variety of namespaces and other
metadata for annotations
that will be used throughout the ontology.  Of particular
interest is the version string towards the end.  As of
3 July 2024, version 4.5 has been published.  We
are working with version 4.2.

ontoProc uses an interface (via the reticulate package)
to the 
[owlready2](https://owlready2.readthedocs.io/en/v0.42/) python modules
distributed via pypi.

Ingestion is typically fast and returns an instance of 
the S3 class `owlents`.
```{r doop}
orde = setup_entities(pa)
orde
```

The most common conceptual relationship encoded in the
ontologies to be considered is "is-a".  Given any
term in the ontology, its ancestors are terms
that are successively more general.

Let's look at two terms.  Given an owlents instance
`x` and a numeric vector `n`, `x[n]` 'subsets' the
information to the selected terms.
```{r lkanc}
orde[1000:1001]
ancestors(orde[1000:1001])
labels(orde[1000:1001])
```
Here we are seeing a mix of R and python references.
As this package matures, this mixing should diminish.

We can use character tags to subset owlents
instances as well.  This is how we "decode" two
of the ancestors of `orde[1000]`.

```{r deco1}
labels(orde[c("Orphanet_377788", "Orphanet_557493")])
```

A new feature of ontoProc's OWL processing is the
availability of searching labels with regular expression
matching.  This feature is built in to owlready2.

```{r lksrch}
lsrch = search_labels(orde$owlfn, "*lipodystrophy*")
length(lsrch)
```
To visualize relationships among the terms,
a plot method leverages the ontologyIndex
packages from CRAN.

```{r lkpl}
plot(orde, names(lsrch)[15:25])
```

## Cell ontology

The `owl2cache` function retrieves OWL from a URL or file
and places it in a cache to avoid repetitious retrievals.  The
default cache is the one defined by `BiocFileCache::BiocFileCache()`.
Here we work with the cell ontology.  `setup_entities` will use
owlready2 python modules to parse the OWL and produce an
instance of S3 class `owlents`.

```{r getcl,message=FALSE}
library(ontoProc)
clont_path = owl2cache(url="http://purl.obolibrary.org/obo/cl.owl")
# need to fix!
tmp = readLines(clont_path)
bad = grep("STATO_0000416", tmp)[1:2]  # see https://github.com/obophenotype/cell-ontology/issues/3237
tmp = tmp[-bad]
bad = grep("STATO_0000663", tmp)[1:2]  # see https://github.com/obophenotype/cell-ontology/issues/3237
tmp = tmp[-bad]
tf = tempfile()
writeLines(tmp, tf)
cle = setup_entities(tf)
cle
```

A plot method is available.  Given a vector of tags
as reported in OWL (no colons are used), the plot method produces
an ontologyIndex instance and runs onto_plot2
on the result.

```{r lkcl}
sel = c("CL_0000492", "CL_0001054", "CL_0000236", 
"CL_0000625", "CL_0000576", 
"CL_0000623", "CL_0000451", "CL_0000556")
plot(cle, sel)
```

## Illustrations with Human Phenotype ontology

We'll obtain and ad hoc selection of
15 UBERON term names and visualize
the hierarchy.

```{r gethp}
hpont_path = owl2cache(url="http://purl.obolibrary.org/obo/hp.owl")
hpents = setup_entities(hpont_path)
kp = grep("UBER", hpents$clnames, value=TRUE)[21:35]
plot(hpents, kp)
```

The prefixes of class names in the ontology
give a sense of its scope.
```{r lkta}
t(t(table(sapply(strsplit(hpents$clnames, "_"), "[", 1))))
```

To characterize human phenotypes ontologically, 
CL, GO, CHEBI, and
UBERON play significant roles.

# Caching

We described `owl2cache` above.  By default, BiocFileCache's cache
is used.  Query it for all available OWL files.

```{r lkbcc}
library(BiocFileCache)
ca = BiocFileCache()
bfcquery(ca, "owl")
```

At present there is no attempt to identify ontology version.
This enhancement may be introduced in a future revision of ontoProc.

Once we have the path, `setup_entities` will deserialize the content to the
`owlents` class.

# Ontology of GWAS outcomes in the GWAS catalog

In this section we examine a package that is under development,
[gwasCatSearch](https://github.com/ccb-hms/gwasCatSearch).  This uses
ideas from the [semantic sql project](https://github.com/INCATools/semantic-sql)
to improve performance of the most common operations on ontologies.

We'll use the search app at [shinyapps.io](https://vjcitn.shinyapps.io/gwasCatSearch)
to explore connections between the GWAS catalog and the Experimental Factor
Ontology used to annotate GWAS outcomes.  Intermediate components are
corpustools for enhancing discovery in natural language formulations of
phenotypes, and SQLite, with tables representing ontology terms and
their ancestors and descendants.

This segment gives some sense of how outcomes can be sorted
according to the frequency of recorded hits in the catalog,
and the associated EFO terms decoded.
```{r lksem}
library(gwasCatSearch)
library(RSQLite)
library(dplyr)
con = gwasCatSearch_dbconn()
dbListTables(con)
tbl(con, "gwascatalog_associations") |> select(MAPPED_TRAIT_URI) |> 
    group_by(MAPPED_TRAIT_URI) |> summarise(n=n()) |> arrange(desc(n))
tbl(con, "efo_labels") |> filter(Subject %in% c("EFO:0004340", "EFO:0004612"))
```
